Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-08-26 manu-draft.qmd
adding lulc map

The health of estuarine systems and coastal habitats is inextricably linked to land uses and management of the watershed [@Yoskowitz15]. Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. Pervasive stressors can negatively impact the services provided by these habitats. Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Related changes in habitats in response to climate change (sea-level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].

@@

Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (Fig 1). Its watershed encompasses another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. It is subtropical, and within the current (2020) ecotone for mangrove and salt marsh habitats. The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural systems within and surrounding Tampa Bay, including direct removal of habitat (including dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

>>>>

The health of estuarine systems and coastal habitats is inextricably linked to land uses and management of the watershed [@Yoskowitz15]. Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. Pervasive stressors can negatively impact the services provided by these habitats. Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Related changes in habitats in response to climate change (sea-level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].  Management of coastal habitats requires an understanding of how stressores impact these resources, while balancing competing societal uses.

@@

Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap). The watershed includes another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. The climate is subtropical and within the current (2020) ecotone for mangrove and salt marsh habitats. The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

![Land use and land cover for the Tampa Bay watershed, Florida, USA.](figs/lulcmap.png){#fig-lulcmap fig-align="center"}
